---
title: Introduction to R!
author: Chris Kypridemos, Vincy Huang
date: October 2020
output: 
  learnr::tutorial:
    id: "week-3"
    version: 0.2
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: "Welcome to learnr tutorials!"
---

```{=html}
<style type="text/css">
  body{
  font-size: 12pt;
}
</style>
```
```{r setup, include=FALSE}
library(learnr)
library(submitr)
library(googlesheets4)
library(gradethis)
library(CKteachR)
knitr::opts_chunk$set(echo = FALSE)
learnr::tutorial_options(
  exercise.timelimit = 60, 
  exercise.checker = 
  gradethis::grade_learnr)
heights_cm <- c(184, 162, 145, 200, 178, 154, 172, 142)
heights_m <- heights_cm / 100 # convert cm to m
```

```{r}
submitr::login_controls()
```

```{r context = "server", echo = FALSE}
options(tutorial.storage = "none")

out <-
  setup_progress_monitoring(
    rstudioapi::getActiveProject(),
    "PHDL1_week3",
    "15oXGGq0fgFL7kN3ZCcouhJhnIY_wJCDj3tPu6nyug-g",
    "122eUZo4Q3VJ3hTy9ySoOTITPfNkrb1-p4nNeQm78w5I", # week 3 spreadsheet
    "UoL.MPH.datalab@gmail.com"
  )

submitr::shiny_logic(input, output, session, out$vfun, out$storage_actions)
```

<!--End of submitr setup-->

## R as a calculator

### Remember to login!

Using your name and your password in the relevant boxes at the top of the page 

### Simple maths
You can use R as a calculator for example you can type `2 + 2` and get `4`

```{r calculator-addition1, echo=TRUE}
2 + 2
```

Use the box below to calculate the product of 6 and 7. 

```{r calculator-multiplication1, exercise=TRUE, exercise.cap="Graded", exercise.lines=7}


```

<div id="calculator-multiplication1-hint">
**Hint:** We need to calculate 6 times 7. Therefore `6 * 7` is the correct answer. 
</div>

```{r calculator-multiplication1-solution}
6 * 7
```

```{r calculator-multiplication1-check}
gradethis::grade_code(correct = "6 * 7")
```

### Using powers

- That's all good but how can I use powers? I see no way to enter a superscript. i.e. 2^2^

- You can use the special symbol `^` to denote the power. I.e `2^2` means 2^2^

- Remember that a square root is the inverse operation to the power of 2. So to calculate $\sqrt{2}$ in R you can simply type `2^(1/2)`. Note the use of brackets! `2^1/2` gives the wrong results because R first calculates `2^1` and then divised by `2`.

- Another option to denote a square root is the function `sqrt()`. I.e. `sqrt(2)`

### Using mathematical expressions

- So now can you use R to calculate the result of $((6 * 7) + 4 - 7) / 2^8$?
```{r zrydcewguzcaojrr, exercise=TRUE, exercise.cap="Graded", exercise.lines=7}


```

<div id="zrydcewguzcaojrr-hint">
**Hint:** All you need is to type the expression from the question into the box. 
</div>

```{r zrydcewguzcaojrr-check}
# check code
gradethis::grade_result(
  fail_if( ~ !all.equal(.result,
                        ((6 * 7) + 4 - 7) / 2^8),
           gradethis::random_encourage()),
  pass_if(~ TRUE, gradethis::random_praise()),
  glue_correct = "{.message}"
)
```


### Other 'numbers' to look out for

- R uses some special 'numbers' to represent infinity or 'undefined'

- Now, try to calculate 1/0, -1/0, and 0/0 and find out what is the result

```{r caculator-other, exercise = TRUE, exercise.lines=7}


```

## Assignment

### Assignment operators

-   We can tell R to remember things so that we can call them later
-   We do this using either assignment operators `=` or `<-` to assign a value to a variable. I.e. `var <- 42` creates a variable with name 'var' that holds the value 42.

```{r assignment-operator, echo=FALSE}
learnr::question(
  "Which of these are assignment operations? (Select ALL that apply)",
  answer('x == 6', correct = FALSE),
  answer('x <- 6', correct = TRUE),
  answer('x = 6',  correct = TRUE),
  incorrect = "Give it another try. We use `=` or `<-` as assignment operators",
  random_answer_order = TRUE,
  allow_retry = TRUE,
  options = list(foobar = "Flag")
)
```

### Things to remember when naming variables

- Naming variables is a hard task. Ideally, variable names should be short to type easily, but informative so your code is more legible

- Remember R is CASE SENSITIVE! `John` is not the same as `john`  

- Variable names cannot begin with numbers or include spaces

- Try assign 9 to a variable named 3john and see what happens

```{r var-name, exercise = TRUE, exercise.lines=3}

```

### Variables task

What do you think the following lines of R code will output at the end?

```{r echo=TRUE}
x <- 2
x <- x + 5 
```

```{r assignment-test, echo=FALSE}
learnr::question("What is the value of x?",
         answer("x5", correct = FALSE),
         answer("7", correct = TRUE, message = gradethis::random_praise()),
         answer("0",  correct = FALSE),
         allow_retry = TRUE,
         random_answer_order = TRUE
)
```

## Vectors

We often have more than a single number, which we combine into a *vector* using
the function `c()`, e.g.

```{r vectors, echo=TRUE}
c(184, 162, 145, 200, 178, 154, 172, 142)
```
As before with single values, we can assign the vector to a variable

```{r vectors2, echo=TRUE}
heights_cm <- c(184, 162, 145, 200, 178, 154, 172, 142)
heights_cm 
```

Many operations in R are *vectorised*, meaning they apply to all elements of a vector, e.g.

```{r vectors3, echo=TRUE}
heights_m <- heights_cm / 100 # convert cm to m
heights_m 
```
Notice that every element of the vector was divided by 100!

## Selection

### Syntax

-   We might want to select the 5-th value from a vector
-   We use square brackets for this, e.g.

```{r vectors4, echo=TRUE}
heights_m[5]
```
- What happens when we request an element beyond the length of the vector? I.e. `heights_m` has a length of 8. What will happen if we request the 10-th element that doesn't really exists?

```{r vectors5, echo=TRUE}
heights_m[10]
```

- Note R returns `NA` which denotes a missing value

### Selection task 1

```{r selection-quiz}
learnr::question(
  "How do you select  the 7-th value from the vector heights?",
  answer("heights(7)", correct = FALSE, message = gradethis::random_encourage()),
  answer("heights[7]", correct = TRUE, message = gradethis::random_praise()),
  answer("heights([7])",  correct = FALSE, message = gradethis::random_encourage()),
  incorrect = "Give it another try. We use square brackets to select from a vector",
  random_answer_order = TRUE,
  allow_retry = TRUE,
  options = list(foobar = "Flag")
)

```

### Selection task 2

How to select 1-st, 3-rd and 5-th element from the vector heights_m?

```{r bcxvsdeyonyvhjsf,  exercise=TRUE, exercise.cap="Graded", exercise.lines=7}


```

<div id="bcxvsdeyonyvhjsf-hint">
**Hint:** Remember that the 1-st, 3-rd and 5-th element is the vector `c(1, 3, 5)` 
</div>

```{r bcxvsdeyonyvhjsf-solution}
# solution code
heights_m[c(1, 3, 5)]
```

```{r bcxvsdeyonyvhjsf-check}
# check code
gradethis::grade_code()
```

## Logic

### Simple logical statements

Boils down to something being `TRUE` or `FALSE`\
`x > y`  asks: is x greater than y?\
`x < y`  asks: is x less than y?\
`x == y` asks: is x equal to y?\
`x >= y` asks: is x greater than or equal to y\
`x <= y` asks: is x less than or equal to y?\

Note the different meaning of `=` and `==`!!! The first assigns, the second checks equality

Basic examples:

```{r logic-1, echo = TRUE}
5 < 10
3 > 5
```

### More logic operators

There are three useful operators that help us combine simple logical statements\
`&` means 'AND'. It returns `TRUE` only if the expression in both sides is `TRUE`\
`|` means 'OR'. It returns `TRUE` if any of the expression in both sides is `TRUE`\
`!` means 'NOT'. It converts `TRUE` to `FALSE` and vice versa\

Using these symbols we can combine logical statements, for example:

```{r logic-combine, echo=TRUE}
(5 < 10) & (3 < 5)
(1 > 2) | (4 > 3)
!(5 < 10)
5 != 10 # same as !(5 == 10)
```

### Logic task
```{r logic-combine1-quiz}
learnr::question(
  "Evaluate the logical statement `(1 < 2) & (10 < 9)`",
  answer("TRUE", correct = FALSE, message = gradethis::random_encourage()),
  answer("FALSE", correct = TRUE, message = gradethis::random_praise()),
  incorrect = "Not quite right...",
  random_answer_order = TRUE,
  allow_retry = TRUE,
  options = list(foobar = "Flag")
)
```

```{r logic-combine2-quiz}
learnr::question(
  "Evaluate the logical statement `(1 < 2) | (10 < 9)`",
  answer("FALSE", correct = FALSE, message = gradethis::random_encourage()),
  answer("TRUE", correct = TRUE, message = gradethis::random_praise()),
  incorrect = "Not quite right...",
  random_answer_order = TRUE,
  allow_retry = TRUE,
  options = list(foobar = "Flag")
)

```

```{r logic-combine3-quiz}
learnr::question(
  "Evaluate the logical statement `!(1 < 2)`",
  answer("TRUE", correct = FALSE, message = gradethis::random_encourage()),
  answer("FALSE", correct = TRUE, message = gradethis::random_praise()),
  incorrect = "Not quite right...",
  random_answer_order = TRUE,
  allow_retry = TRUE,
  options = list(foobar = "Flag")
)

```


### Logic & selection
We can use a *logical vector* to pick out elements of a vector. Let's try first with a
logical vector that is the same length as the vector we want to partition.

```{r, echo=TRUE}
heights_cm # to remind us its content
logic_vec <- c(TRUE, FALSE, TRUE, FALSE, TRUE, FALSE, TRUE, FALSE)
heights_cm[logic_vec]
```

We can simplify this further. Notice from above that our goal is to get every other element of the vector. When R uses vectorised operations and the vectors have unequal lengths, it silently *recycles* the shorter vector(s) to the length of the longest vector. For example:

```{r, echo=TRUE}
logic_vec2 <- c(TRUE, FALSE) 
heights_cm[logic_vec2] # same as before because of recycling
```
R silently recycles `logic_vec2` internaly to reach the length of `heights_cm` by repeating `logic_vec2`. So `logic_vec2` silently becomes `c(TRUE, FALSE, TRUE, FALSE, TRUE, FALSE, TRUE, FALSE)` when interacts with `heights_cm`.

### Logic & selection (c'nued)

How to extract heights greater than 160cm?

Remember that `heights_cm > 160` returns a logical vector

```{r echo=TRUE}
ind <- heights_cm > 160
ind
```

So then we can simply do
```{r echo=TRUE}
heights_cm[ind]
```

Or we can combine the 3 lines above into 1

```{r echo=TRUE}
heights_cm[heights_cm > 160]
```


### Logic & selection task

Can you extract heights less than or equal to 180cm in one line?

```{r tllsfpaflolywmon, exercise=TRUE, exercise.cap="Graded", exercise.lines=1}


```

<div id="tllsfpaflolywmon-hint">
**Hint:** Try to modify the code from the previous example `heights_cm[heights_cm > 160]` 
</div>

```{r tllsfpaflolywmon-solution}
heights_cm[heights_cm <= 180]
```

```{r tllsfpaflolywmon-check}
# check code
gradethis::grade_code()
```

### Remember to submit your answers!

Click the submit button at the top of the page
